---
title: "Group rows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{group-rows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette introduces nice and easy way to display grouped data frame created by `dplyr::group_by`.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(ftExtra)
library(dplyr)
```

```{r setup2, ref.label='setup', eval=FALSE}

```


```{r}
grouped_iris <- iris %>%
  group_by(Species) %>%
  slice(1, 2)

grouped_mtcars <- mtcars %>%
  mutate(model = rownames(mtcars)) %>%
  head %>%
  select(model, cyl, mpg, disp, am) %>%
  group_by(am, cyl)
```

# Groups as titles

## Single grouping columns

```{r}
grouped_iris %>% as_flextable()
```


```{r}
grouped_iris %>% as_flextable(hide_grouplabel = TRUE)
```

## Multiple grouping columns

```{r}
grouped_mtcars %>% as_flextable()
```

# Groups as merged columns

By specifying `as_flextable(groups_to = 'merged')`,
grouping variables are merged vertically.
In this case, the default theme is changed to `flextable::theme_vanilla`
because the booktab theme is not intuitive.

## Single grouping variable

```{r}
grouped_iris %>%
  as_flextable(groups_to = 'merged')
```

## Multiple grouping variables

```{r}
grouped_mtcars %>%
  as_flextable(groups_to = 'merged') %>%
  flextable::theme_vanilla()
```

## Position of grouping variables

Grouping variables are moved to left by default.
If you want to keep their positions, specify `group_pos = 'asis'`.

```{r}
grouped_mtcars %>%
  as_flextable(groups_to = 'merged', groups_pos = 'asis') %>%
  flextable::theme_vanilla()
```
