---
title: "Create multi level headers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create multi level headers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette introduces how to create multi level headers.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = "dplyr" %in% rownames(installed.packages())
)
```

```{r setup}
library(ftExtra)
```

# Prepare flextable

```{r}
ft <- iris[1:2, ] %>% flextable
ft
```

# Split headers

The `split_header()` function generates multi-level headers by separating original headers (row names) by delimiters.

```{r}
ft %>% split_header()
```

The `sep` argument defines delimiters by regular expression.
The default delimiters are `.` and `_`.
Let's see what happens when `seop = "e"`

```{r}
ft %>% split_header(sep = "e")
```

# Span headers

The `span_header()` function also generates multi-level headers.
Unlike the `split_header()` function, the `span_header()` function merges adjacent headers if they have same values.

```{r}
ft %>% span_header()
```

